---
title: "02_mitogenomes"
output: github_document
date: "2025-11-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(ggtree)
library(RColorBrewer)
library(viridis)
library(scales)
library(tidyverse)
library(ggpubr)
library(dplyr)
library(purrr)
library(readr)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(ggrepel)
library(scales)
library(igraph)
library(visNetwork)
library(ape)
library(pegas)
library(dplyr)
```

```{r}
sample_meta <- read.csv("Dugong_metadata_all_incl_Tian_paper_final.csv", header = TRUE)
sample_ids <- read_lines("02_Mitogenomes_local/sample_ids.txt")

library(writexl)

write_xlsx(sample_meta, "sample_meta.xlsx")


# Filter metadata to only those IDs and select the trait
traits <- sample_meta %>%
  filter(bioplatforms_library_id %in% sample_ids) %>%
  select(ID = bioplatforms_library_id, Region = location_text)

traits <- traits %>%
  slice(match(sample_ids, ID))

# Write to CSV
write_csv(traits, "traits.csv")
```


We aligned ALL the files to the mitochondrial genome which is from the same animal as the nuclear genome we used as a reference.

Ran scripts 00.stagein.sh to get all files from Acacia.
Then 01.fix_bam_headers.sh, 02.sort_bams.sh and 03.call_hapl_consensus.sh.
Next up, we converted to fasta using 04.convert_to_fasta.sh.
Then I downloaded goalign and ran 05.goalign.sh

I used an online tool to convert fasta to nexus format.

Then I ran 07.run_iqtree.sh and downloaded the resulting .contree file in here.

```{r, echo = TRUE}
tree <- read.tree("02_Mitogenomes_local//AllHapsClean_all.fasta.contree")
# Root the tree using sample "628131" from Palau
rooted_tree <- root(tree, outgroup = "628131", resolve.root = TRUE)

# Plot it
plot(rooted_tree, show.tip.label = TRUE)

# Save to file
write.tree(rooted_tree, file = "02_Mitogenomes_local//AllHapsClean_rooted.tree")
```


```{r}
meta <- read.csv("traits.csv", colClasses = c("ID" = "character"))

# Join tree tip labels with metadata using sample IDs
tree_data <- data.frame(label = tree$tip.label) %>%
  left_join(sample_meta, by = c("label" = "bioplatforms_library_id"))

detailed_tree_data <- data.frame(label = tree$tip.label) %>%
  left_join(sample_meta, by = c("label" = "bioplatforms_library_id"))

traits <- sample_meta %>%
  filter(bioplatforms_library_id %in% sample_ids) %>%
  select(ID = bioplatforms_library_id, Region = location_text)
```

```{r}
west_to_east_order <- c(
  "Shark Bay","Exmouth Gulf","Barrow Island","Port Hedland","Broome",
  "One Arm Point","Darwin","Pellew Islands","Wellesley Islands","Torres Strait",
  "Turtle Head Island","Starcke","Elim Beach","Daintree","Port Douglas",
  "Yarrabah","Hinchinbrook","Bowling Green Bay","Airlie Beach","Clairview",
  "Shoalwater Bay","Emu Park","Gladstone","Hervey Bay","Great Sandy Strait",
  "Moreton Bay","Arcadia Vale","Sydney"
)

my_location_colors_without_n <- c(
  "Shark Bay"="#30123BFF","Exmouth Gulf"="#3C3286FF","Barrow Island"="#4452C0FF",
  "Port Hedland"="#476FE7FF","Broome"="lightskyblue","One Arm Point"="lightblue",
  "Darwin"="paleturquoise","Pellew Islands"="#18D8C9FF","Wellesley Islands"="darkcyan",
  "Torres Strait"="darkgreen","Turtle Head Island"="seagreen","Starcke"="darkseagreen",
  "Elim Beach"="lightgreen","Daintree"="greenyellow","Port Douglas"="yellow",
  "Yarrabah"="gold","Hinchinbrook"="orange","Bowling Green Bay"="darkorange",
  "Airlie Beach"="coral","Clairview"="tomato","Shoalwater Bay"="indianred",
  "Emu Park"="firebrick","Gladstone"="darkred","Hervey Bay"="maroon",
  "Great Sandy Strait"="darkmagenta","Moreton Bay"="orchid",
  "Arcadia Vale"="plum","Sydney"="pink"
)


filled_shapes <- c(21, 22, 23, 24, 25)
loc_shapes <- setNames(rep(filled_shapes, length.out = length(west_to_east_order)),
                       west_to_east_order)

my_location_colors <- c(
  "Shark Bay (n=16)"         = "#30123BFF",
  "Exmouth Gulf (n=14)"      = "#3C3286FF",
  "Barrow Island (n=2)"      = "#4452C0FF",
  "Port Hedland (n=4)"       = "#476FE7FF",
  "Broome (n=7)"             = "lightskyblue",
  "One Arm Point (n=2)"      = "lightblue",
  "Darwin (n=6)"             = "paleturquoise",
  "Pellew Islands (n=11)"    = "#18D8C9FF",
  "Wellesley Islands (n=3)"  = "darkcyan",
  "Torres Strait (n=12)"     = "darkgreen",
  "Turtle Head Island (n=2)" = "seagreen",
  "Starcke (n=1)"            = "darkseagreen",
  "Elim Beach (n=1)"         = "lightgreen",
  "Daintree (n=1)"           = "greenyellow", 
  "Port Douglas (n=1)"       = "yellow",
  "Yarrabah (n=2)"           = "gold",
  "Hinchinbrook (n=1)"       = "orange",
  "Bowling Green Bay (n=4)"  = "darkorange",
  "Airlie Beach (n=3)"       = "coral",
  "Clairview (n=9)"          = "tomato",
  "Shoalwater Bay (n=7)"     = "indianred",
  "Emu Park (n=1)"           = "firebrick",
  "Gladstone (n=4)"          = "darkred",
  "Hervey Bay (n=24)"        = "maroon",
  "Great Sandy Strait (n=23)"= "darkmagenta",
  "Moreton Bay (n=43)"       = "orchid",
  "Arcadia Vale (n=1)"       = "plum",
  "Sydney (n=1)"             = "pink"
)    

shape_names <- c(
  "circle" = 21,
  "square" = 22,
  "diamond" = 23,
  "triangle_up" = 24,
  "triangle_down" = 25
)

loc_shapes_named <- c(
  "Shark Bay" = shape_names["circle"],
  "Exmouth Gulf" = shape_names["square"],
  "Barrow Island" = shape_names["diamond"],
  "Port Hedland" = shape_names["triangle_up"],
  "Broome" = shape_names["triangle_down"],
  "One Arm Point" = shape_names["circle"],
  "Darwin" = shape_names["square"],
  "Pellew Islands" = shape_names["diamond"],
  "Wellesley Islands" = shape_names["triangle_up"],
  "Torres Strait" = shape_names["triangle_down"],
  "Turtle Head Island" = shape_names["circle"],
  "Starcke" = shape_names["square"],
  "Elim Beach" = shape_names["diamond"],
  "Daintree" = shape_names["triangle_up"],
  "Port Douglas" = shape_names["triangle_down"],
  "Yarrabah" = shape_names["circle"],
  "Hinchenbrook" = shape_names["square"],
  "Borlwing Green Bay" = shape_names["diamond"],
  "Airlie Beach" = shape_names["triangle_up"],
  "Clairview" = shape_names["triangle_down"],
  "Shoalwater Bay" = shape_names["circle"],
  "Emu Park" = shape_names["square"],
  "Gladstone" = shape_names["diamond"],
  "Hervey Bay" = shape_names["triangle_up"],
  "Great Sandy Strait" = shape_names["triangle_down"],
  "Moreton Bay" = shape_names["circle"],
  "Arcadia Vale" = shape_names["square"],
  "Sydney" = shape_names["diamond"]
)

# Set fill colors and filled shapes
loc_colors <- my_location_colors  
```

```{r}

detailed_tree_data$location_text <- factor(
  detailed_tree_data$location_text,
  levels = west_to_east_order
)

loc_colors <- setNames(turbo(length(west_to_east_order)), west_to_east_order)

scale_color_manual(
  values = my_location_colors_without_n,
  breaks = west_to_east_order 
)

loc_colors["Abu Dhabi"] <- "grey70"
loc_colors["Palau"] <- "grey30"
loc_colors["New Caledonia"] <- "lightgrey"
loc_colors["Malaysia"] <- "darkgrey"
loc_colors["Okinawa, Japan"] <- "green"

extra_locs <- c("Malaysia", "Okinawa, Japan", "Abu Dhabi", "Palau", "New Caledonia")
extra_shapes <- rep(filled_shapes, length.out = length(extra_locs))
loc_shapes <- c(loc_shapes, setNames(extra_shapes, extra_locs))

p <- ggtree(tree) %<+% tree_data +
  geom_tippoint(aes(color = location_text, shape =location_text), size = 2) +
  geom_tiplab(aes(label = label), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n, name = "Region") +
  scale_shape_manual(values = loc_shapes, name = "Region") +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.8, "cm")
  )

print(p)

tree_data <- tree_data %>%
  mutate(label_location = paste0(label, " (", location_text, ")"))

# Plot with the combined label
p_label <- ggtree(tree) %<+% tree_data +
  geom_tippoint(aes(color = location_text, shape = location_text), size = 2) +
  geom_tiplab(aes(label = label_location), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n, name = "Region") +
  scale_shape_manual(values = loc_shapes, name = "Region") +
  theme(
    legend.position = "none",  
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.8, "cm")
  )

print(p_label)
```


Focusing only on Australia now:
```{r}
# Define only Australian locations
australian_locs <- c(
  "Shark Bay","Exmouth Gulf","Barrow Island","Port Hedland","Broome",
  "One Arm Point","Darwin","Pellew Islands","Wellesley Islands","Torres Strait",
  "Turtle Head Island","Starcke","Elim Beach","Daintree","Port Douglas",
  "Yarrabah","Hinchinbrook","Bowling Green Bay","Airlie Beach","Clairview",
  "Shoalwater Bay","Emu Park","Gladstone","Hervey Bay","Great Sandy Strait",
  "Moreton Bay","Arcadia Vale","Sydney"
)

tree_data_aus <- tree_data %>%
  filter(location_text %in% australian_locs) %>%
  mutate(label_location = paste(label, location_text, sep = "_"))

# Drop non-Australian tips from the tree
tree_aus <- drop.tip(tree, tree$tip.label[!tree$tip.label %in% tree_data_aus$label])

# Plot
aus_label <- ggtree(tree_aus) %<+% tree_data_aus +
  geom_tippoint(aes(color = location_text, shape = location_text), size = 2) +
  geom_tiplab(aes(label = label_location), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n[australian_locs]) +
  scale_shape_manual(values = loc_shapes[australian_locs]) +
  theme(legend.position = "none")  # removes the legend

print(aus_label)
```

**NO CONTROL REGION**

I first downloaded seqkit:
wget https://github.com/shenwei356/seqkit/releases/download/v2.3.1/seqkit_linux_amd64.tar.gz


Then I had to index the reference
```{python, eval=FALSE, python.reticulate = FALSE}
samtools faidx mDugDug1.MT.20230221.fasta
```

Then I had to adjust a lot of things, check scripts 08 - 11

```{r}
tree_no_control <- read.tree("02_Mitogenomes_local/AllMito_noCRClean.fasta.contree")
# Root the tree using sample "628131"
rooted_tree <- root(tree_no_control, outgroup = "628131", resolve.root = TRUE)

# Plot it
plot(rooted_tree, show.tip.label = TRUE)

# Save to file
write.tree(rooted_tree, file = "02_Mitogenomes_local/AllMito_noCRClean_rooted.tree")

# Join tree tip labels with metadata using sample IDs
tree_no_control_data <- data.frame(label = tree_no_control$tip.label) %>%
  left_join(sample_meta, by = c("label" = "bioplatforms_library_id"))

detailed_tree_no_control_data <- data.frame(label = tree_no_control$tip.label) %>%
  left_join(sample_meta, by = c("label" = "bioplatforms_library_id"))

traits <- sample_meta %>%
  filter(bioplatforms_library_id %in% sample_ids) %>%
  select(ID = bioplatforms_library_id, Region = location_text)
```

```{r}
tree_no_control_data_aus <- tree_no_control_data %>%
  filter(location_text %in% australian_locs) %>%
  mutate(label_location = paste(label, location_text, sep = "_"))

# Drop non-Australian tips from the tree
tree_no_control_aus <- drop.tip(rooted_tree, rooted_tree$tip.label[!rooted_tree$tip.label %in% tree_no_control_data_aus$label])

# Plot
aus_label <- ggtree(tree_no_control_aus) %<+% tree_no_control_data_aus +
  geom_tippoint(aes(color = location_text, shape = location_text), size = 2) +
  geom_tiplab(aes(label = label_location), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n[australian_locs]) +
  scale_shape_manual(values = loc_shapes[australian_locs]) +
  theme(legend.position = "none") +
  geom_treescale(x = 0, y = -7, width = 0.10) 

print(aus_label)

```

Now, I'm gonna try to run a tree with the full sequences.
I am using the online mafft tool to make an alignment.

```{r}
tree_full_algined <- read.tree("02_Mitogenomes_local/AllMito_full_mafft_algined.fasta.contree")
# Root the tree using sample "628131"
rooted_tree_full_aligned <- root(tree_full_algined, outgroup = "628131", resolve.root = TRUE)

# Join tree tip labels with metadata using sample IDs
tree_full_algined_data <- data.frame(label = tree_full_algined$tip.label) %>%
  left_join(sample_meta, by = c("label" = "bioplatforms_library_id"))

traits <- sample_meta %>%
  filter(bioplatforms_library_id %in% sample_ids) %>%
  select(ID = bioplatforms_library_id, Region = location_text)

tree_full_algined_data_aus <- tree_full_algined_data %>%
  filter(location_text %in% australian_locs) %>%
  mutate(label_location = paste(label, location_text, sep = "_"))

# Drop non-Australian tips from the tree
tree_full_aligned_aus <- drop.tip(
  rooted_tree_full_aligned,
  rooted_tree_full_aligned$tip.label[!rooted_tree_full_aligned$tip.label %in% tree_full_algined_data_aus$label]
)

# Plot
aus_full_aligned <- ggtree(tree_full_aligned_aus) %<+% tree_full_algined_data_aus +
  geom_tippoint(aes(color = location_text, shape = location_text), size = 2) +
  geom_tiplab(aes(label = label_location), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n[australian_locs]) +
  scale_shape_manual(values = loc_shapes[australian_locs]) +
  theme(legend.position = "none")

print(aus_full_aligned)

# Drop sample 681998 from the tree
tree_full_aligned_aus_pruned <- drop.tip(tree_full_aligned_aus, "681998")

# Also remove it from the metadata
tree_full_aligned_data_aus_pruned <- tree_full_algined_data_aus %>%
  filter(label != "681998")

# Plot
aus_full_aligned_pruned <- ggtree(tree_full_aligned_aus_pruned) %<+% tree_full_aligned_data_aus_pruned +
  geom_tippoint(aes(color = location_text, shape = location_text), size = 2) +
  geom_tiplab(aes(label = label_location), size = 2, color = "black") +
  scale_color_manual(values = my_location_colors_without_n[australian_locs]) +
  scale_shape_manual(values = loc_shapes[australian_locs]) +
  theme(legend.position = "none") +
  geom_treescale(x = 0, y = -7, width = 0.01) 

print(aus_full_aligned_pruned)

```





**ADDING ONLINE SEQUENCES**

Ok so I downloaded all the available sequences on NCBI that have the words "Dugong" and "control" in them as a fasta file and the metadata as a .gb file.
Then I opened them in Geneious, aligned everything to the referene genome, including all the sequences from me and the Tian et al. paper, and then visually extracted the part that ALL sequences have in common.

I cleaned up the files like this on Pawsey:
```{python, eval=FALSE}
sed -E 's/^>([A-Za-z0-9_.-]+).*/>\1/' Extracted_region_from_all_files.fasta > Extracted_region_from_all_files_clean.fasta

sed -E 's/^>([^ ]+?)_assembled_to.*$/>\1/' Extracted_region_from_all_files_clean.fasta > Extracted_region_from_all_files_clean_simple.fasta
```

And then ran iqtree 
```{python, eval=FALSE}
iqtree2 -s Extracted_region_from_all_files_clean_simple.fasta -m TEST -bb 1000 -alrt 1000 -nt AUTO
```

```{r}
library(stringr)
library(dplyr)
library(purrr)
library(tibble)

# Path to combined GenBank file
gb_file <- "02_Mitogenomes_local/sequence.gb"

# Read the full file
lines <- readLines(gb_file)

# Split into individual GenBank records
records <- str_split(paste(lines, collapse = "\n"), "//")[[1]]

# Function to extract metadata from a record
extract_gb_record <- function(record) {
  isolate <- str_match(record, '/isolate="([^"]+)"')[,2]
  geo <- str_match(record, '/geo_loc_name="([^"]+)"')[,2]
  country <- str_match(record, '/country="([^"]+)"')[,2]
  organism <- str_match(record, '/organism="([^"]+)"')[,2]
  accession <- str_match(record, 'ACCESSION\\s+([A-Z0-9_]+)')[,2]
  
  # Pick first non-missing location
  region <- coalesce(geo[!is.na(geo)][1],
                     country[!is.na(country)][1],
                     NA_character_)
  
  # Assign "Thailand" to any isolate starting with "DU" if region missing
  if (!is.na(isolate[1]) && str_detect(isolate[1], "^DU") && (is.na(region) || region == "")) {
    region <- "Thailand"
  }
  
  tibble(
    ID = isolate[!is.na(isolate)][1],
    Region = region,
    Organism = organism[!is.na(organism)][1],
    Accession = accession[!is.na(accession)][1],
    Source = "NCBI"
  )
}

# Apply to all records
ncbi_meta <- map_dfr(records, extract_gb_record) %>%
  filter(!is.na(ID))

# Inspect results
print(ncbi_meta, n = 10)


```


```{r}
library(dplyr)
library(readr)

# Load local metadata
sample_meta <- read_csv("02_Mitogenomes_local/traits.csv",
                        col_types = cols(ID = col_character()))

# Select only the relevant columns and add a Source column
local_meta <- sample_meta %>%
  transmute(
    ID = ID,
    Region = Region,
    Source = "Local"
  )

# Prepare NCBI metadata (from previous extraction)
ncbi_meta_simple <- ncbi_meta %>%
  transmute(
    ID = ID,
    Region = Region,
    Source = "NCBI"
  )

# Combine both sources
all_meta <- bind_rows(local_meta, ncbi_meta_simple)

# Remove duplicates if any
all_meta <- all_meta %>% distinct(ID, .keep_all = TRUE)

head(all_meta)

```

```{r}
# Load consensus tree
NCBI_control_region_tree <- read.tree("02_Mitogenomes_local/Extracted_region_from_all_files_clean_simple.fasta.contree")

# Create a dataframe with tip labels
tree_data <- tibble(label = NCBI_control_region_tree$tip.label)

# Join metadata by sample ID
tree_data <- tree_data %>%
  left_join(all_meta, by = c("label" = "ID"))

head(tree_data)

p <- ggtree(NCBI_control_region_tree) %<+% tree_data +
  geom_tiplab(aes(label = paste0(label, " (", Region, ")")),
              size = 2.5, align = TRUE) +
  theme_tree2() +
  ggtitle("Dugong Control Region Phylogeny")

p
```

Trying again with a more cleaned-up fasta file from geneious, removing any sequences that had gaps in the beginning or end of the sequence.

```{python, eval=FALSE}
sed -E 's/^>([A-Za-z0-9_.-]+).*/>\1/' Control_region_NCBI_cleaned.fasta > Control_region_NCBI_cleaned_IDs.fasta.fasta

sed -E 's/^>([^ ]+?)_assembled_to.*$/>\1/' Control_region_NCBI_cleaned_IDs.fasta.fasta > Control_region_NCBI_cleaned_IDs_simple.fasta
```

And then ran iqtree again
```{python, eval=FALSE}
iqtree2 -s Control_region_NCBI_cleaned_IDs_simple.fasta -m TEST -bb 1000 -alrt 1000 -nt AUTO
```

```{r}
extract_gb_record <- function(record) {
  isolate <- str_match(record, '/isolate="([^"]+)"')[,2]
  geo <- str_match(record, '/geo_loc_name="([^"]+)"')[,2]
  country <- str_match(record, '/country="([^"]+)"')[,2]
  organism <- str_match(record, '/organism="([^"]+)"')[,2]
  accession <- str_match(record, 'ACCESSION\\s+([A-Z0-9_]+)')[,2]
  
  region <- coalesce(
    geo[!is.na(geo)][1],
    country[!is.na(country)][1],
    NA_character_
  )

  # Assign Thailand based on isolate prefix DU*
  if (!is.na(isolate[1]) && str_detect(isolate[1], "^DU") && (is.na(region) || region == "")) {
    region <- "Thailand"
  }
  
  tibble(
    ID = accession,          # <<< KEY CHANGE — use accession as the matching ID
    Isolate = isolate,
    Region = region,
    Organism = organism[!is.na(organism)][1],
    Accession = accession,
    Source = "NCBI"
  )
}


ncbi_meta <- map_dfr(records, extract_gb_record)

all_meta <- bind_rows(
    local_meta,
    ncbi_meta
)

tree_data <- tibble(label = Control_final_maybe$tip.label)

tree_data <- tree_data %>%
  left_join(all_meta, by = c("label" = "ID"))

p_final <- ggtree(Control_final_maybe) %<+% tree_data +
  geom_tiplab(aes(label = paste0(label, " (", Region, ")")), 
              size = 2.3, align = TRUE) +
  theme_tree2() +
  ggtitle("Dugong Control Region Phylogeny")

p_final

traits <- tree_data %>%
  select(ID = label, Region) %>% 
  arrange(ID)

write.csv(traits, "traits.csv", row.names = FALSE)

# Read CSV safely
df <- read.csv("traits.csv", stringsAsFactors = FALSE, check.names = FALSE)

# Remove quotes and trim whitespace
df$Region <- gsub('^"|"$', '', df$Region)
df$Region <- trimws(df$Region)

# Replace spaces, colons, and commas with underscores
df$Region <- gsub("[ ,:]", "_", df$Region)

# Optional: replace multiple consecutive underscores with a single one
df$Region <- gsub("_+", "_", df$Region)

# Write cleaned CSV for fasta2nexus.py
write.csv(df, "traits_all_clean.csv", row.names = FALSE, quote = FALSE)
```

With this traits file and the fasta file, I made a nexus file!
Then I also made a traits file.
```{r}
traits_all_clean <- read.csv("02_Mitogenomes_local/traits_all_clean.csv", header = TRUE)

trait_matrix <- traits_all_clean %>%
  mutate(value = 1) %>% 
  pivot_wider(names_from = Region, values_from = value, values_fill = 0)

trait_matrix

out_file <- "traits_block.txt"
con <- file(out_file, open = "w", encoding = "UTF-8")

locations <- colnames(trait_matrix)[-1]  # all Region columns

cat("BEGIN TRAITS;\n", file = con)
cat("Dimensions NTRAITS=", length(locations), ";\n", sep = "", file = con)
cat("Format labels=yes missing=? separator=Comma;\n", file = con)
cat("TraitLabels ", paste(locations, collapse = " "), ";\n", sep = "", file = con)
cat("Matrix\n", file = con)

apply(trait_matrix, 1, function(row) {
  cat(row[1], " ", paste(row[-1], collapse=","), "\n", sep="", file=con)
})

cat(";\nend;\n", file = con)
close(con)

cat("Wrote file to: ", normalizePath(out_file), "\n")
```
I copied this part into the nexus file, that then worked in popart!

```{r}
# Load NEXUS alignment
nexus <- read.nexus.data("02_Mitogenomes_local/Control_region_NCBI_cleaned_IDs_simple.nexus")   

# Convert to a DNAbin object
dna_bin <- as.DNAbin(nexus)

# Compute haplotypes
haps <- haplotype(dna_bin)

# Display haplotype network
net <- haploNet(haps)

```






Now I'm trying to do diversity stats.

```{r}
Mito_aligned <- read.dna("02_Mitogenomes_local/AllMito_full_mafft_algined.fasta", format="fasta")

meta <- read.csv("traits.csv", colClasses = c("ID" = "character"))
```


```{r}
hap <- haplotype(Mito_aligned) 
haplo_div <- hap.div(Mito_aligned) 
nuc_diversity <- nuc.div(Mito_aligned) 

hap
haplo_div
nuc_diversity
```

Now I'm rerunning everything in groups

```{r}
# Read DNA and metadata
meta <- read.csv("traits.csv", colClasses = c("ID" = "character"))

meta$Region <- trimws(meta$Region)

# Define clusters
meta_clusters <- meta %>%
  mutate(Cluster = case_when(
    Region %in% c("Shark Bay") ~ "Shark Bay",
    Region %in% c("Exmouth Gulf", "Broome", "One Arm Point", "Port Hedland", "Barrow Island") ~ "Kimberley",
    Region %in% c("Pellew Islands", "Wellesley Islands") ~ "Gulf of Carpentaria",
    Region %in% c("Torres Strait", "Turtle Head Island", "Starcke", "Elim Beach", "Daintree", "Port Douglas", "Yarrabah", "Hinchinbrook", "Bowling Green Bay", "Airlie Beach") ~ "Northern Queensland",
    Region %in% c("Clairview", "Shoalwater Bay", "Emu Park", "Gladstone", "Hervey Bay", "Great Sandy Strait", "Moreton Bay") ~ "Southern Queensland",
    Region %in% c("Arcadia Vale", "Sydney") ~ "Sydney",
    TRUE ~ Region
  ))

# Match metadata to DNA
meta2 <- meta_clusters %>% filter(ID %in% labels(Mito_aligned))
meta2 <- meta2[match(labels(Mito_aligned), meta2$ID), ]

meta2_clean <- meta2[!is.na(meta2$Cluster), ]
Mito_aligned_clean <- Mito_aligned[!is.na(meta2$Cluster), , drop = FALSE]

# Compute diversity per cluster
clusters <- unique(meta2$Cluster)

clusters <- unique(meta2_clean$Cluster)

diversity_by_cluster <- lapply(clusters, function(cl) {
  inds <- which(meta2_clean$Cluster == cl)
  subdna <- Mito_aligned_clean[inds, , drop = FALSE]  

  hap_div <- if(length(inds) == 1) NA else hap.div(subdna)
  nuc_div <- if(length(inds) == 1) NA else nuc.div(subdna)
  hap_count <- length(haplotype(subdna))

  data.frame(
    Cluster = cl,
    N = length(inds),
    Haplotypes = hap_count,
    Haplotype_diversity = hap_div,
    Nucleotide_diversity = nuc_div
  )
})

div_df <- do.call(rbind, diversity_by_cluster)
div_df

library(ape)
library(dplyr)
library(pegas)

# 1. Read your data
Mito_aligned <- read.dna("02_Mitogenomes_local/AllMito_full_mafft_algined.fasta", format="fasta")
meta <- read.csv("traits.csv", colClasses = c("ID" = "character"))

# 2. Match metadata to your sequences
meta <- meta %>%
  filter(ID %in% labels(Mito_aligned)) %>%
  mutate(Region = trimws(Region)) %>%
  mutate(Cluster = case_when(
    Region %in% c("Shark Bay") ~ "Shark Bay",
    Region %in% c("Exmouth Gulf", "Broome", "One Arm Point", "Port Hedland", "Barrow Island") ~ "Kimberley",
    Region %in% c("Pellew Islands", "Wellesley Islands") ~ "Gulf of Carpentaria",
    Region %in% c("Torres Strait", "Turtle Head Island", "Starcke", "Elim Beach", "Daintree", 
                  "Port Douglas", "Yarrabah", "Hinchinbrook", "Bowling Green Bay", "Airlie Beach") ~ "Northern Queensland",
    Region %in% c("Clairview", "Shoalwater Bay", "Emu Park", "Gladstone", "Hervey Bay", 
                  "Great Sandy Strait", "Moreton Bay") ~ "Southern Queensland",
    Region %in% c("Arcadia Vale", "Sydney") ~ "Sydney",
    TRUE ~ Region
  ))

# Ensure sequence order matches metadata
meta <- meta[match(labels(Mito_aligned), meta$ID), ]

# 3. Compute diversity per cluster
clusters <- unique(meta$Cluster)

diversity_by_cluster <- lapply(clusters, function(cl) {
  
  inds <- which(meta$Cluster == cl)
  subdna <- Mito_aligned[inds]
  
  # Ensure DNAbin is always correctly formatted
  if(length(subdna) == 1) subdna <- as.DNAbin(subdna)
  
  # Compute haplotype and nucleotide diversity if more than 1 sequence
  hap_div <- if(length(inds) > 1) hap.div(subdna) else NA
  nuc_div <- if(length(inds) > 1) nuc.div(subdna) else NA
  hap_count <- length(haplotype(subdna))
  
  data.frame(
    Cluster = cl,
    N = length(inds),
    Haplotypes = hap_count,
    Haplotype_diversity = hap_div,
    Nucleotide_diversity = nuc_div
  )
})

# Combine results into a single table
diversity_table <- do.call(rbind, diversity_by_cluster)

# 4. Save results
write.csv(diversity_table, "mito_diversity_by_cluster.csv", row.names = FALSE)

# View table
diversity_table


```
I downloaded all sequences with these search terms from NCBI:
"Dugong dugon"[Organism] AND (mitochondrion OR "control region" OR "D-loop") 

Aligned them with Geneious to the Dugdug mt genome from the same animal as the nuclear reference genome, and cut out the part of the sequence that all of them had in common (removing a few sequences that didn't fit at all, or were way shorter).


```{r}
# Path to combined GenBank file
gb_file <- "02_Mitogenomes_local/All-ncbi-downloads-metadata.gb"

# Read the full file
lines <- readLines(gb_file)

# Split into individual GenBank records
records <- str_split(paste(lines, collapse = "\n"), "//")[[1]]

# Function to extract metadata from a record
extract_gb_record <- function(record) {
  isolate <- str_match(record, '/isolate="([^"]+)"')[,2]
  geo <- str_match(record, '/geo_loc_name="([^"]+)"')[,2]
  country <- str_match(record, '/country="([^"]+)"')[,2]
  organism <- str_match(record, '/organism="([^"]+)"')[,2]
  accession <- str_match(record, 'ACCESSION\\s+([A-Z0-9_]+)')[,2]
  
  # Pick first non-missing location
  region <- coalesce(geo[!is.na(geo)][1],
                     country[!is.na(country)][1],
                     NA_character_)
  
  # Assign "Thailand" to any isolate starting with "DU" if region missing
  if (!is.na(isolate[1]) && str_detect(isolate[1], "^DU") && (is.na(region) || region == "")) {
    region <- "Thailand"
  }
  
  tibble(
    ID = isolate[!is.na(isolate)][1],
    Region = region,
    Organism = organism[!is.na(organism)][1],
    Accession = accession[!is.na(accession)][1],
    Source = "NCBI"
  )
}

# Apply to all records
ncbi_meta <- map_dfr(records, extract_gb_record) %>%
  filter(!is.na(ID))

# Inspect results
print(ncbi_meta, n = 10)

ncbi_meta %>%
  filter(is.na(Region))

```

