---
title: "Dugong mitochondrial genomes"
output: github_document
date: "23/04/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(ggplot2)
library(stringr)
library(svglite)
library(grid)
library(ape)
library(adegenet)
library(poppr)
library(ggtree)
library(tidyverse)
library(pegas)
library(ggrepel)
library(maps)
library(poppr)
library(gridExtra)
library(cowplot)
library(ggtext)
library(openxlsx)

all_location_cols <- c(
 "Shark Bay"         = "#30123BFF",
  "Exmouth Gulf"      = "#3C3286FF",
  "Barrow Island"      = "#4452C0FF",
  "Port Hedland"       = "#476FE7FF",
  "Broome"             = "lightskyblue",
  "One Arm Point"      = "lightblue",
  "Darwin"             = "paleturquoise",
  "Pellew Islands"    = "#18D8C9FF",
  "Wellesley Islands"  = "darkcyan",
  "Torres Strait"     = "darkgreen",
  "Turtle Head Island" = "seagreen",
  "Starcke"            = "darkseagreen",
  "Elim Beach"         = "lightgreen",
  "Daintree"           = "greenyellow", 
  "Port Douglas"       = "yellow",
  "Yarrabah"           = "gold",
  "Hinchinbrook"       = "orange",
  "Townsville"  = "darkorange",
  "Airlie Beach"       = "coral",
  "Clairview"          = "tomato",
  "Shoalwater Bay"     = "indianred",
  "Emu Park"           = "firebrick",
  "Gladstone"          = "darkred",
  "Hervey Bay"        = "maroon",
  "Great Sandy Strait"= "darkmagenta",
  "Moreton Bay"       = "orchid",
  "Sydney"             = "pink",
  "UAE"             = "rosybrown4",
  "Palau"         = "mediumblue",
  "Malaysia"      = "green",
  "Japan"         = "lightpink1",
  "New Caledonia" = "khaki1"
)



all_locations_order <- c(
  # Australia
  "Shark Bay", "Exmouth Gulf", "Barrow Island", "Port Hedland", "Broome",
  "One Arm Point", "Darwin", "Pellew Islands", "Wellesley Islands",
  "Torres Strait", "Turtle Head Island", "Starcke", "Elim Beach", "Daintree",
  "Port Douglas", "Yarrabah", "Hinchinbrook", "Townsville",
  "Airlie Beach", "Clairview", "Shoalwater Bay", "Emu Park", "Gladstone",
  "Hervey Bay", "Great Sandy Strait", "Moreton Bay", "Sydney",

  # Outside Australia
  "UAE", "Palau", "Malaysia", "Japan", "New Caledonia"
)
```

Reads were mapped to the reference genome using bwa mem. This included my samples, Tian et al.'s data (https://doi.org/10.1038/s41467-024-49769-x), and individual samples DRR251525 and ERR5621402 from NCBI.

```{bash, eval = FALSE}
ref=mDugDug1.MT.20230221.fasta

bams=(mapped_marked_bams/*.bam)

input=${bams[$SLURM_ARRAY_TASK_ID]}

bn=$(basename $input)

output=mitobams/${bn%_marked.bam}_mito.bam

mkdir -p mitobams

echo "Starting mapping for ${output}"

samtools collate -Oun128 ${input} | samtools fastq - \
  | bwa mem -pt8 ${ref} - | samtools view -F 4 -O BAM - > ${output} 

echo "Done ${output}"
```

To get fasta files for each one, I ran this command:
```{bash, eval = FALSE}
for bam in mitobams/*_mito_fixed_sorted.bam; do
    sample=$(basename "$bam" "_mito_fixed_sorted.bam")
    echo "Processing $sample"

    # Generate VCF
    bcftools mpileup -f mDugDug1.MT.20230221.fasta "$bam" | \
        bcftools call -mv -Oz -o "mito_fasta/${sample}.vcf.gz"

    # Index VCF
    bcftools index "mito_fasta/${sample}.vcf.gz"

    # Generate consensus FASTA
    bcftools consensus -f mDugDug1.MT.20230221.fasta \
        "mito_fasta/${sample}.vcf.gz" > "mito_fasta/${sample}.fasta"
done
```


For comparisons with publicly available control region data, I downloaded all sequences with these search terms from NCBI:
"Dugong dugon"[Organism] AND (mitochondrion OR "control region" OR "D-loop") 

I also have my own samples and the ones from the Tian paper available as a fasta file.

I used Geneious to align the sequences to the mt DugDug sequence that is from the same animal as our nuclear reference genome. I also added a manatee sequence as an outgroup. I went through the sequences and made sure they all align ok and that one out-of-place base does not cause a huge mismatch. The final extract was 

Then I cut out only the part that aligns for all these sequences, which obviously seems to be the control region, but it is admittedly only a very small part of that region still (304bp).

```{r}
gb_file <- "02_Mitogenomes_local/All-ncbi-downloads-metadata.gb"

lines <- readLines(gb_file)

records <- str_split(paste(lines, collapse = "\n"), "//")[[1]]

# Function to extract metadata from a record
extract_gb_record <- function(record) {
  isolate   <- str_match(record, '/isolate="([^"]+)"')[,2]
  geo       <- str_match(record, '/geo_loc_name="([^"]+)"')[,2]
  country   <- str_match(record, '/country="([^"]+)"')[,2]
  organism  <- str_match(record, '/organism="([^"]+)"')[,2]
  accession <- str_match(record, 'ACCESSION\\s+([A-Z0-9_]+)')[,2]
  version   <- str_match(record, 'VERSION\\s+([A-Z0-9_.]+)')[,2]

  region <- coalesce(geo[!is.na(geo)][1],
                     country[!is.na(country)][1],
                     NA_character_)

  if (!is.na(isolate[1]) && str_detect(isolate[1], "^DU") && (is.na(region) || region == "")) {
    region <- "Thailand"
  }

  tibble(
    TreeID  = version[!is.na(version)][1],   
    ID      = isolate[!is.na(isolate)][1],
    Region  = region,
    Organism = organism[!is.na(organism)][1],
    Accession = accession[!is.na(accession)][1],
    Source = "NCBI"
  )
}

ncbi_meta <- map_dfr(records, extract_gb_record)

ncbi_meta_simple <- ncbi_meta %>% 
  select(TreeID, Region, Source)

# Load master metadata
dugong_meta <- read_csv(
  "Dugong_metadata_FINAL.csv",
  col_types = cols(.default = col_character())
)

dugong_meta <- dugong_meta %>%
  mutate(
    decimal_longitude_private = as.numeric(decimal_longitude_private),
    decimal_latitude_private  = as.numeric(decimal_latitude_private)
  )

dugong_meta <- dugong_meta %>%
  mutate(
    country = dplyr::recode(
      country,
      "New_Caledonia" = "New Caledonia"
    )
  )

traits_from_master <- dugong_meta %>%
  transmute(
    TreeID = bioplatforms_library_id,
    Region = location_text,
    Source = "Local"
  ) %>%
  filter(!is.na(TreeID))


# Combine LOCAL + NCBI
all_meta_full <- bind_rows(traits_from_master, ncbi_meta_simple) %>%
  distinct(TreeID, .keep_all = TRUE)

all_meta_full <- bind_rows(traits_from_master, ncbi_meta_simple) %>%
  distinct(TreeID, .keep_all = TRUE) %>%
  mutate(
    Region = if_else(
      TreeID %in% c("ON227090.1", "ON227091.1"),
      "New Caledonia",
      Region
    )
  )


write_csv(
  traits_from_master %>% select(TreeID, Region),
  "02_Mitogenomes_local/traits_local_final.csv"
)

Control_region_tree <- read.tree("02_Mitogenomes_local/Extract8.fasta.contree")

Control_region_tree_rooted <- root(
  Control_region_tree,
  outgroup = "AM904728.1",
  resolve.root = TRUE
)

tree_data <- tibble(
  label = Control_region_tree_rooted$tip.label
) %>%
  left_join(all_meta_full, by = c("label" = "TreeID"))

```

Next, I'm turning this into a popart-useable nexus file to make Haplotype networks.

```{r}
region_map <- c(
  # WA
  "Shark_Bay" = "WA",
  "Exmouth_Gulf" = "WA",
  "Broome" = "WA",
  "Port_Hedland" = "WA",
  "One_Arm_Point" = "WA",
  "Barrow_Island" = "WA",

  # QLD
  "Starcke" = "QLD",
  "Elim_Beach" = "QLD",
  "Torres_Strait" = "QLD",
  "Hinchinbrook" = "QLD",
  "Yarrabah" = "QLD",
  "Turtle_Head_Island" = "QLD",
  "Emu_Park" = "QLD",
  "Hervey_Bay" = "QLD",
  "Clairview" = "QLD",
  "Gladstone" = "QLD",
  "Daintree" = "QLD",
  "Port_Douglas" = "QLD",
  "Great_Sandy_Strait" = "QLD",
  "Bowling_Green_Bay" = "QLD",
  "Townsville" = "QLD",
  "Airlie_Beach" = "QLD",
  "Shoalwater_Bay" = "QLD",
  "Moreton_Bay" = "QLD",

  # NT
  "Darwin" = "NT",
  "Wellesley_Islands" = "NT",
  "Pellew_Islands" = "NT",

  # NSW
  "Arcadia_Vale" = "NSW",

  # International
  "United_Arab_Emirates" = "UAE",
  "United_Arab_Emirates__Abu_Dhabi" = "UAE",
  "Abu_Dhabi" = "UAE",

  "Tanzania" = "Tanzania",
  "Tanzania__Rufiji" = "Tanzania",
  "Tanzania__Rufiji_Mafia_District" = "Tanzania",

  "Indian_Ocean__Nicobar_Islands" = "Nicobar_Islands",
  "Mauritius__Fort_Frederik_Hendrik" = "Mauritius"
)

traits_for_popart <- tree_data %>%
  select(label, Region) %>%
  distinct() %>%
  rename(ID = label) %>%
  mutate(

    Region_original = Region,

    Region = gsub("[^A-Za-z0-9_]", "_", Region),

    Region = case_when(

      # Existing labels
      Region %in% c("QLD", "WA", "NT", "NSW",
                    "Thailand", "UAE",
                    "Tanzania", "Mauritius",
                    "Palau", "Malaysia",
                    "Indonesia", "India",
                    "Japan", "Comoros",
                    "Kenya", "Egypt",
                    "Sudan", "Yemen",
                    "Djibouti", "Bahrain",
                    "Mozambique",
                    "Madagascar",
                    "Sri_Lanka",
                    "Philippines",
                    "New_Caledonia",
                    "Papua_New_Guinea",
                    "Nicobar_Islands",
                    "Indian_Ocean",
                    "Indian_Ocean__Africa",
                    "Indian_Ocean__Red_Sea",
                    "Australia") ~ Region,

      # Queensland
      grepl("Queensland|Torres_Strait|Townsville|Moreton_Bay|Hervey_Bay|Gladstone|Clairview|Yarrabah|Daintree|Port_Douglas|Shoalwater_Bay|Airlie_Beach|Bowling_Green_Bay|Hinchinbrook|Elim_Beach|Starcke|Emu_Park|Magnetic_Island|Bamaga",
             Region) ~ "QLD",

      # WA
      grepl("Western_Australia|Shark_Bay|Exmouth_Gulf|Broome|Port_Hedland|One_Arm_Point|Barrow_Island",
             Region) ~ "WA",

      # NT
      grepl("Darwin|Wellesley_Islands|Pellew_Islands",
             Region) ~ "NT",

      # NSW
      grepl("Sydney|Arcadia_Vale",
             Region) ~ "NSW",

      # UAE
      grepl("United_Arab_Emirates|Abu_Dhabi",
             Region) ~ "UAE",

      TRUE ~ Region
    )
  )

traits_for_popart <- traits_for_popart %>%
  mutate(
    Region = ifelse(
      grepl("^Thailand", Region),
      "Thailand",
      Region
    )
  )

traits_for_popart <- traits_for_popart %>%
  mutate(
    Region = case_when(

      Region %in% c(
        "Great_Sandy_Strait",
        "Turtle_Head_Island"
      ) ~ "QLD",

      Region %in% c(
        "Tanzania__Rufiji",
        "Tanzania__Rufiji_Mafia_District"
      ) ~ "Tanzania",

      Region == "Mauritius__Fort_Frederik_Hendrik" ~ "Mauritius",

      TRUE ~ Region
    )
  )

trait_matrix <- traits_for_popart %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = Region,
    values_from = value,
    values_fill = 0
  )

locations <- colnames(trait_matrix)[-1]

lines <- c(
  "BEGIN TRAITS;",
  paste0("Dimensions NTRAITS=", length(locations), ";"),
  "Format labels=yes missing=? separator=Comma;",
  paste0("TraitLabels ", paste(locations, collapse = " "), ";"),
  "Matrix",
  apply(trait_matrix, 1, function(row) {
    paste0(row[1], " ", paste(row[-1], collapse = ","))
  }),
  ";",
  "END;"
)

#writeLines(lines, "02_Mitogenomes_local/traits_block.txt")
```

Now I'm trying to make my life easier by making a list of each haplogroup and which locations it's from.


```{r}
hap_raw <- read.delim(
  "02_Mitogenomes_local/Extract8_log_final_maybe.log",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

hap_with_region <- hap_raw %>%
  left_join(
    traits_for_popart %>% select(ID, Region),
    by = c("Matching.Sequences" = "ID")
  )

restricted_ids <- c(
  "SRR29328970",
  "SRR29328990",
  "628133",
  "KJ944384.1",
  "EU835771.1",
  "SRR29329011",
  "SRR29329056",
  "SRR29329036",
  "DugDug_mtGenome",
  "628099",
  "628119"
)

widespread_ids <- c(
  "628129",
  "628076",
  "MT136741.1",
  "MH704428.1",
  "628101",
  "SRR29329065",
  "682056",
  "794820",
  "628124",
  "628083",
  "628135",
  "628073",
  "681980",
  "628075",
  "628105",
  "628060"
)

hap_with_region <- hap_with_region %>%
  mutate(
    lineage_range = case_when(
      Matching.Sequences %in% restricted_ids ~ "restricted",
      Matching.Sequences %in% widespread_ids ~ "widespread",
      TRUE ~ NA_character_
    )
  )
```

And here's some statistics on these. I clicked on "Statistics" and then "Identical sequences" in popart and wrote the output to a log file.

```{r}
haplotype_network_seqs <- read.dna(
  "02_Mitogenomes_local/Extract8.fasta",
  format = "fasta"
)

# Keeping only regions with >= 6 samples
valid_regions <- traits_for_popart %>%
  dplyr::count(Region) %>%
  dplyr::filter(n >= 6) %>%
  dplyr::pull(Region)

metadata_filtered <- traits_for_popart %>%
  filter(Region %in% valid_regions)


calc_stats <- function(dna_subset) {

  n <- nrow(dna_subset)

  haps <- haplotype(dna_subset)

  Nh <- nrow(haps)

  h <- hap.div(dna_subset)

  S <- length(seg.sites(dna_subset))

  pi <- nuc.div(dna_subset)

  if(S > 0) {
    taj <- tajima.test(dna_subset)

    D <- round(as.numeric(taj[[1]]), 3)
    p <- round(taj$Pval.normal, 3)
  } else {
    D <- NA
    p <- NA
  }

  data.frame(
    n = n,
    Nh = Nh,
    h = round(h, 3),
    S = S,
    pi = round(pi, 5),
    tajimas_D = D,
    tajima_p = p
  )
}

# Calculate stats per region
results <- list()

for(region in unique(metadata_filtered$Region)) {

  ids <- metadata_filtered$ID[
    metadata_filtered$Region == region
  ]

  matched_ids <- intersect(
    ids,
    rownames(haplotype_network_seqs)
  )

  # Skip empty groups
  if(length(matched_ids) < 2) {
    next
  }

  # Convert to numeric indices
  keep <- which(
    rownames(haplotype_network_seqs) %in%
      matched_ids
  )

  dna_subset <- haplotype_network_seqs[
    keep,
  ]

  cat(region,
      nrow(dna_subset),
      "\n")

  results[[region]] <- calc_stats(dna_subset)
}

# Combine results
summary_table <- bind_rows(
  results,
  .id = "Region"
)

# Sort by sample size
summary_table <- summary_table %>%
  arrange(desc(n))

summary_table

write.csv(
  summary_table,
  "haplo_diversity_summary.csv",
  row.names = FALSE
)
```


Next up, I'm looking into the full mitochondrial genomes.

```{r}
all_mt <- read.dna("02_Mitogenomes_local/All_incl_ERR_DRR.fasta", format = "fasta")

d <- dist.dna(all_mt, model = "N", pairwise.deletion = TRUE)

pcoa_all <- cmdscale(d, k = 2, eig = TRUE)

eig_vals <- pcoa_all$eig

pos_eigs <- eig_vals[eig_vals > 0]

var_explained <- pos_eigs / sum(pos_eigs) * 100

pc1_pct <- round(var_explained[1], 1)
pc2_pct <- round(var_explained[2], 1)

pc1_pct
pc2_pct

plot(
  pcoa_all$points[,1],
  pcoa_all$points[,2],
  pch = 19,
  xlab = "Axis 1",
  ylab = "Axis 2"
)

text(
  pcoa_all$points[,1],
  pcoa_all$points[,2],
  labels = rownames(pcoa_all$points),
  pos = 3,
  cex = 0.6
)

var_explained <- pcoa_all$eig / sum(pcoa_all$eig) * 100

pc1_lab <- sprintf("Axis 1 (%.1f%%)", var_explained[1])
pc2_lab <- sprintf("Axis 2 (%.1f%%)", var_explained[2])
```

And here's how I calculated the statistics for the full genomes, plus control region only, and everything except for control region.

```{r}
# Haplotype diversity
hap.div(all_mt)

# Nucleotide diversity
nuc.div(all_mt)

nrow(all_mt)

# Number of haplotypes
haps_all_mt <- haplotype(all_mt)
nrow(haps_all_mt)

#Tajima's D
tajima.test(all_mt)

all_mt_control_only <- read.dna("02_Mitogenomes_local/All_incl_ERR_DRR_control_only.fasta", format = "fasta")

# Haplotype diversity
hap.div(all_mt_control_only)

# Nucleotide diversity
nuc.div(all_mt_control_only)

# Number of haplotypes
haps_all_mt_control_only <- haplotype(all_mt_control_only)
nrow(haps_all_mt_control_only)

#Tajima's D
tajima.test(all_mt_control_only)

all_mt_no_control<- read.dna("02_Mitogenomes_local/All_incl_ERR_DRR_no_control.fasta", format = "fasta")

# Haplotype diversity
hap.div(all_mt_no_control)

# Nucleotide diversity
nuc.div(all_mt_no_control)

# Number of haplotypes
haps_all_mt_no_control <- haplotype(all_mt_no_control)
nrow(haps_all_mt_no_control)

#Tajima's D
tajima.test(all_mt_no_control)
```

Australia only:
```{r}
mt_ids <- rownames(all_mt)
mt_meta <- dugong_meta %>%
  filter(bioplatforms_library_id %in% mt_ids)

new_row <- mt_meta[1, ]
new_row[,] <- NA

new_row$bioplatforms_library_id <- "DugDug_mtGenome"
new_row$country <- "Australia"

mt_meta <- bind_rows(mt_meta, new_row)

all_mt_aus <- all_mt[
  rownames(all_mt) %in%
    mt_meta$bioplatforms_library_id[mt_meta$country == "Australia"],
]

# Sample size
nrow(all_mt_aus)

# Haplotype diversity
hap.div(all_mt_aus)

# Nucleotide diversity
nuc.div(all_mt_aus)

# Number of haplotypes
haps_aus <- haplotype(all_mt_aus)
nrow(haps_aus)

# Tajima's D
tajima.test(all_mt_aus)
```

I then split my data into widespread and restricted group within Australia (excluding New Caledonian samples!).

```{r}
widespread_group <- read.dna("02_Mitogenomes_local/65_samples_widespread.fasta", format = "fasta")

widespread_group <- widespread_group[
  rownames(widespread_group) != "DugDug_mtGenome",
]

# Check sample size
nrow(widespread_group)

# Haplotype diversity
hap.div(widespread_group)

# Nucleotide diversity
nuc.div(widespread_group)

# Number of haplotypes
haps_widespread_group <- haplotype(widespread_group)
nrow(haps_widespread_group)

#Tajima's D
tajima.test(widespread_group)

restricted_group <- read.dna("02_Mitogenomes_local/86_samples_restricted.fasta", format = "fasta")

# Haplotype diversity
hap.div(restricted_group)

# Nucleotide diversity
nuc.div(restricted_group)

# Number of haplotypes
haps_restricted_group <- haplotype(restricted_group)
nrow(haps_restricted_group)

#Tajima's D
tajima.test(restricted_group)
```

And then just to check, ran a PCOA on the separate groups.

```{r}
dist_wide <- dist.dna(
  widespread_group,
  model = "TN93",   #based on iqtree
  pairwise.deletion = TRUE
)

pcoa_wide <- pcoa(dist_wide)

pcoa_df <- data.frame(
  Sample = rownames(pcoa_wide$vectors),
  PC1 = pcoa_wide$vectors[,1],
  PC2 = pcoa_wide$vectors[,2]
)

pcoa_df_wide <- data.frame(
  bioplatforms_library_id = rownames(pcoa_wide$vectors),
  PC1 = pcoa_wide$vectors[, 1],
  PC2 = pcoa_wide$vectors[, 2],
  stringsAsFactors = FALSE
)

pcoa_df_wide <- pcoa_df_wide %>%
  left_join(
    dugong_meta %>%
      select(bioplatforms_library_id, location_text, country),
    by = "bioplatforms_library_id"
  )

pcoa_df_wide <- pcoa_df_wide %>%
  mutate(
    location_text = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Moreton Bay",
      location_text
    ),
    country = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Australia",
      country
    )
  )

ggplot(pcoa_df_wide, aes(PC1, PC2, colour = location_text)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_colour_manual(values = all_location_cols, drop = FALSE) +
  theme_classic() +
  labs(
    x = paste0(
      "PCoA 1 (",
      round(pcoa_wide$values$Relative_eig[1] * 100, 1),
      "%)"
    ),
    y = paste0(
      "PCoA 2 (",
      round(pcoa_wide$values$Relative_eig[2] * 100, 1),
      "%)"
    ),
    colour = "Location"
  )
```

```{r}
dist_restr <- dist.dna(
  restricted_group,
  model = "F84",   #based on iqtree it's HKY, but this seems the closest of the available models
  pairwise.deletion = TRUE
)

pcoa_restr <- pcoa(dist_restr)

pcoa_df_restr <- data.frame(
  Sample = rownames(pcoa_restr$vectors),
  PC1 = pcoa_restr$vectors[,1],
  PC2 = pcoa_restr$vectors[,2]
)


pcoa_df_restr <- data.frame(
  bioplatforms_library_id = rownames(pcoa_restr$vectors),
  PC1 = pcoa_restr$vectors[, 1],
  PC2 = pcoa_restr$vectors[, 2],
  PC3 = pcoa_restr$vectors[, 3],
  stringsAsFactors = FALSE
)

pcoa_df_restr <- pcoa_df_restr %>%
  left_join(
    dugong_meta %>%
      select(bioplatforms_library_id, location_text, country),
    by = "bioplatforms_library_id"
  )

pcoa_df_restr <- pcoa_df_restr %>%
  mutate(
    location_text = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Moreton Bay",
      location_text
    ),
    country = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Australia",
      country
    )
  )

ggplot(pcoa_df_restr, aes(PC1, PC2, colour = location_text)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_colour_manual(values = all_location_cols, drop = FALSE) +
  theme_classic() +
  labs(
    x = paste0(
      "PCoA 1 (",
      round(pcoa_restr$values$Relative_eig[1] * 100, 1),
      "%)"
    ),
    y = paste0(
      "PCoA 2 (",
      round(pcoa_restr$values$Relative_eig[2] * 100, 1),
      "%)"
    ),
    colour = "Location"
  )

ggplot(pcoa_df_restr, aes(PC2, PC3, colour = location_text)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_colour_manual(values = all_location_cols, drop = FALSE) +
  theme_classic() +
  labs(
    x = paste0(
      "PCoA 2 (",
      round(pcoa_restr$values$Relative_eig[2] * 100, 1),
      "%)"
    ),
    y = paste0(
      "PCoA 3 (",
      round(pcoa_restr$values$Relative_eig[3] * 100, 1),
      "%)"
    ),
    colour = "Location"
  )
```

```{r}
pcoa_df_all <- data.frame(
  bioplatforms_library_id = rownames(pcoa_all$points),
  PC1 = pcoa_all$points[, 1],
  PC2 = pcoa_all$points[, 2],
  stringsAsFactors = FALSE
) %>%
  left_join(
    dugong_meta %>%
      select(bioplatforms_library_id, location_text, country),
    by = "bioplatforms_library_id"
  ) %>%
  mutate(
    # Fix reference sample
    location_text = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Moreton Bay",
      location_text
    ),
    country = ifelse(
      bioplatforms_library_id == "DugDug_mtGenome",
      "Australia",
      country
    ),

    # FINAL plotting variable 
    plot_location = case_when(
      country == "Australia" ~ location_text,
      country == "UAE" ~ "UAE",
      country == "Palau" ~ "Palau",
      country == "Malaysia" ~ "Malaysia",
      country == "Japan" ~ "Japan",
      country == "New Caledonia" ~ "New Caledonia",
      TRUE ~ NA_character_
    )
  )

pcoa_df_all$plot_location <- factor(
  pcoa_df_all$plot_location,
  levels = names(all_location_cols)
)

non_aus <- c("UAE","Palau","Malaysia","Japan","New Caledonia")

shape_values <- setNames(
  ifelse(names(all_location_cols) %in% non_aus, 15, 16),
  names(all_location_cols)
)

pcoa_aus <- pcoa_df_all %>%
  filter(country == "Australia")

pcoa_non_aus <- pcoa_df_all %>%
  filter(country != "Australia")

final_plot <- ggplot() +
  ## Australian samples: circles
  geom_point(
    data = pcoa_aus,
    aes(PC1, PC2, colour = plot_location),
    size = 3,
    shape = 16
  ) +
  ## Non-Australian samples: squares
  geom_point(
    data = pcoa_non_aus,
    aes(PC1, PC2, colour = plot_location),
    size = 3,
    shape = 15
  ) +
  scale_colour_manual(values = all_location_cols, drop = FALSE) +
  theme_classic() +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    colour = "Location"
  )

final_plot

d <- dist.dna(all_mt, model = "N", pairwise.deletion = TRUE)

pcoa_full <- cmdscale(d, k = 2, eig = TRUE)

pcoa_df_full <- data.frame(
  bioplatforms_library_id = rownames(pcoa_full$points),
  PC1 = pcoa_full$points[, 1],
  PC2 = pcoa_full$points[, 2],
  stringsAsFactors = FALSE
)

#ggsave(
#  filename = "pcoa_plot.png",
#  plot = final_plot,
#  width = 250,
#  height = 160,
#  units = "mm",
#  dpi = 600
#)
```
Then I made a traits file for popart for the whole mitochondrial genome, the mitochondrial genome with just the control region, and the mitochondrial genome without the control region.

```{r}
whole_mt <- read.nexus.data("02_Mitogenomes_local/All_incl_ERR_DRR.nex")

sample_names <- names(whole_mt)

traits_for_popart_full <- data.frame(
  ID = sample_names
) %>%
  left_join(
    dugong_meta %>%
      select(
        bioplatforms_library_id,
        location_text
      ),
    by = c("ID" = "bioplatforms_library_id")
  ) %>%
  rename(
    Region = location_text
  ) %>%
  mutate(
    Region = ifelse(
      ID == "DugDug_mtGenome",
      "Moreton Bay",
      Region
    )
  )

traits_for_popart_full %>%
  filter(is.na(Region))

traits_for_popart_full <- traits_for_popart_full %>%
  mutate(
    Region = gsub(" ", "_", Region)
  )

trait_matrix <- traits_for_popart_full %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = Region,
    values_from = value,
    values_fill = 0
  )



locations <- colnames(trait_matrix)[-1]

out_file <- "traits_block.txt"

con <- file(out_file, open = "w")

cat("BEGIN TRAITS;\n", file = con)

cat(
  "Dimensions NTRAITS=",
  length(locations),
  ";\n",
  sep = "",
  file = con
)

cat(
  "Format labels=yes missing=? separator=Comma;\n",
  file = con
)

cat(
  "TraitLabels ",
  paste(locations, collapse = " "),
  ";\n",
  sep = "",
  file = con
)

cat("Matrix\n", file = con)

apply(trait_matrix, 1, function(row) {

  cat(
    row[1],
    " ",
    paste(row[-1], collapse = ","),
    "\n",
    sep = "",
    file = con
  )

})

cat(";\nEND;\n", file = con)

close(con)
```
And now link those results back to the metadata.
```{r}
popart_edges <- read_tsv(
  "02_Mitogenomes_local/All_incl_ERR_DRR_traits.txt",
  col_names = c("from", "steps", "to")
)

meta_lookup <- dugong_meta %>%
  select(
    bioplatforms_library_id,
    location_text
  ) %>%
  distinct()

annotated_edges <- popart_edges %>%

  left_join(
    meta_lookup,
    by = c("from" = "bioplatforms_library_id")
  ) %>%
  rename(
    from_location = location_text
  ) %>%

  left_join(
    meta_lookup,
    by = c("to" = "bioplatforms_library_id")
  ) %>%
  rename(
    to_location = location_text
  )

annotated_edges

sample_summary <- bind_rows(

  annotated_edges %>%
    select(ID = from, Location = from_location),

  annotated_edges %>%
    select(ID = to, Location = to_location)

) %>%
  filter(!is.na(Location)) %>%
  distinct()

sample_summary
```

Now for the mitochondrial genome without the control region.

```{r}
no_contr <- read.nexus.data("02_Mitogenomes_local/All_incl_ERR_DRR_no_control.nex")

sample_names <- names(no_contr)

traits_for_popart_full <- data.frame(
  ID = sample_names
) %>%
  left_join(
    dugong_meta %>%
      select(
        bioplatforms_library_id,
        location_text
      ),
    by = c("ID" = "bioplatforms_library_id")
  ) %>%
  rename(
    Region = location_text
  ) %>%
  mutate(
    Region = ifelse(
      ID == "DugDug_mtGenome",
      "Moreton Bay",
      Region
    )
  )

traits_for_popart_full %>%
  filter(is.na(Region))

traits_for_popart_full <- traits_for_popart_full %>%
  mutate(
    Region = gsub(" ", "_", Region)
  )

trait_matrix <- traits_for_popart_full %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = Region,
    values_from = value,
    values_fill = 0
  )



locations <- colnames(trait_matrix)[-1]

out_file <- "traits_block_no_control.txt"

con <- file(out_file, open = "w")

cat("BEGIN TRAITS;\n", file = con)

cat(
  "Dimensions NTRAITS=",
  length(locations),
  ";\n",
  sep = "",
  file = con
)

cat(
  "Format labels=yes missing=? separator=Comma;\n",
  file = con
)

cat(
  "TraitLabels ",
  paste(locations, collapse = " "),
  ";\n",
  sep = "",
  file = con
)

cat("Matrix\n", file = con)

apply(trait_matrix, 1, function(row) {

  cat(
    row[1],
    " ",
    paste(row[-1], collapse = ","),
    "\n",
    sep = "",
    file = con
  )

})

cat(";\nEND;\n", file = con)

close(con)
```
Now the same thing just with the control region only.

```{r}
control_region <- read.dna(
  "02_Mitogenomes_local/All_incl_ERR_DRR_control_only.fasta",
  format = "fasta"
)

sample_names <- rownames(control_region)

traits_for_popart_control <- data.frame(
  ID = sample_names
) %>%
  left_join(
    dugong_meta %>%
      select(
        bioplatforms_library_id,
        location_text
      ),
    by = c("ID" = "bioplatforms_library_id")
  ) %>%
  rename(
    Region = location_text
  ) %>%
  mutate(
    Region = ifelse(
      ID == "DugDug_mtGenome",
      "Moreton Bay",
      Region
    )
  )


traits_for_popart_control %>%
  filter(is.na(Region))

traits_for_popart_control <- traits_for_popart_control %>%
  mutate(
    Region = gsub(" ", "_", Region)
  )


trait_matrix <- traits_for_popart_control %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = Region,
    values_from = value,
    values_fill = 0
  )

locations <- colnames(trait_matrix)[-1]

out_file <- "02_Mitogenomes_local/traits_block_control_only.txt"

con <- file(out_file, open = "w")

cat("BEGIN TRAITS;\n", file = con)

cat(
  "Dimensions NTRAITS=",
  length(locations),
  ";\n",
  sep = "",
  file = con
)

cat(
  "Format labels=yes missing=? separator=Comma;\n",
  file = con
)

cat(
  "TraitLabels ",
  paste(locations, collapse = " "),
  ";\n",
  sep = "",
  file = con
)

cat("Matrix\n", file = con)

apply(trait_matrix, 1, function(row) {

  cat(
    row[1],
    " ",
    paste(row[-1], collapse = ","),
    "\n",
    sep = "",
    file = con
  )

})

cat(";\nEND;\n", file = con)

close(con)
```




Now I'm looking at where the differences are along the mitochondrial genome exactly.

```{r}
mt_DNA <- read.dna("02_Mitogenomes_local/All_incl_ERR_DRR.fasta",
                format = "fasta")

bed <- read.table("02_Mitogenomes_local/Dugong_mito_genome.bed",
                  header = FALSE,
                  stringsAsFactors = FALSE)

colnames(bed) <- c("chrom",
                   "start",
                   "end",
                   "gene",
                   "score",
                   "strand")
```


```{r}
seg_sites <- seg.sites(mt_DNA)

length(seg_sites)
head(seg_sites)

snp_df <- data.frame(position = seg_sites)

snp_df$gene <- NA

for(i in 1:nrow(bed)) {

  idx <- snp_df$position >= bed$start[i] &
         snp_df$position <= bed$end[i]

  snp_df$gene[idx] <- bed$gene[i]
}

head(snp_df)

gene_counts <- snp_df %>%
  group_by(gene) %>%
  summarise(n_snps = n()) %>%
  arrange(desc(n_snps))
```

```{r}
window_size <- 200
step_size <- 50

genome_length <- ncol(mt_DNA)

windows <- seq(1,
               genome_length - window_size,
               by = step_size)

snp_counts <- sapply(windows, function(w) {

  sum(seg_sites >= w &
      seg_sites < (w + window_size))
})

plot_df <- data.frame(
  position = windows,
  snps = snp_counts
)

ggplot(plot_df,
       aes(position, snps)) +

  geom_line() +

  theme_bw() +

  labs(x = "Mitogenome position",
       y = "SNPs per 200 bp window")
```

```{r}
samples_widespread <- pcoa_df_full %>%
  filter(PC1 < -40)
samples_restricted <- pcoa_df_full %>%
  filter(PC1 > 40)

restricted_ids <- samples_restricted$bioplatforms_library_id
widespread_ids <- samples_widespread$bioplatforms_library_id

rownames(mt_DNA)[1:10]

sum(restricted_ids %in% rownames(mt_DNA))
sum(widespread_ids %in% rownames(mt_DNA))
```

```{r}
dna_matrix <- as.character(mt_DNA)

restricted_ids <- restricted_ids[
  restricted_ids %in% rownames(dna_matrix)
]

widespread_ids <- widespread_ids[
  widespread_ids %in% rownames(dna_matrix)
]
```


```{r}
fixed_sites <- c()

for(pos in 1:ncol(dna_matrix)) {

  restricted_bases <- unique(
    dna_matrix[restricted_ids, pos]
  )

  widespread_bases <- unique(
    dna_matrix[widespread_ids, pos]
  )

  restricted_bases <- setdiff(
    restricted_bases,
    c("-", "n", "N", "?")
  )

  widespread_bases <- setdiff(
    widespread_bases,
    c("-", "n", "N", "?")
  )

  if(length(restricted_bases) == 1 &
     length(widespread_bases) == 1 &
     restricted_bases[1] != widespread_bases[1]) {

    fixed_sites <- c(fixed_sites, pos)
  }
}

length(fixed_sites)
head(fixed_sites)
```


```{r}
control_region <- data.frame(
  chrom = "Dugong_mito_genome",
  start = 15432,
  end = ncol(dna_matrix),
  gene = "control_region",
  score = 1,
  strand = "+"
)

bed2 <- rbind(bed, control_region)

fixed_df <- data.frame(
  position = fixed_sites
)

fixed_df$gene <- NA

for(i in 1:nrow(bed2)) {

  idx <- fixed_df$position >= bed2$start[i] &
         fixed_df$position <= bed2$end[i]

  fixed_df$gene[idx] <- bed2$gene[i]
}

fixed_summary <- fixed_df %>%
  group_by(gene) %>%
  summarise(n_fixed = n()) %>%
  arrange(desc(n_fixed))

fixed_summary
```

```{r}
window_size <- 100
step_size <- 10

windows <- seq(
  1,
  ncol(dna_matrix) - window_size,
  by = step_size
)

fixed_counts <- sapply(windows, function(w) {

  sum(fixed_sites >= w &
      fixed_sites < (w + window_size))
})

mini_windows <- data.frame(
  start = windows,
  end = windows + window_size,
  fixed_diffs = fixed_counts
)

mini_windows %>%
  arrange(desc(fixed_diffs)) %>%
  head(20)
```

So most of these are in the control region, but some are in the NAD region, which is a coding region!

```{r}
candidate_start <- 12040
candidate_end <- 12170

restricted_consensus <- apply(
  dna_matrix[restricted_ids,
             candidate_start:candidate_end],
  2,
  function(x) {
    names(sort(table(x),
               decreasing = TRUE))[1]
  }
)

widespread_consensus <- apply(
  dna_matrix[widespread_ids,
             candidate_start:candidate_end],
  2,
  function(x) {
    names(sort(table(x),
               decreasing = TRUE))[1]
  }
)

diagnostic_snps <- data.frame(
  position = candidate_start:candidate_end,
  restricted = restricted_consensus,
  widespread = widespread_consensus
)

diagnostic_snps <- diagnostic_snps %>%
  filter(restricted != widespread)

diagnostic_snps
```


```{r}
window_size <- 200
step_size <- 50

windows <- seq(
  1,
  ncol(dna_matrix) - window_size,
  by = step_size
)

fixed_counts <- sapply(windows, function(w) {

  sum(fixed_sites >= w &
      fixed_sites < (w + window_size))
})

fixed_plot <- data.frame(
  position = windows,
  fixed_diffs = fixed_counts
)

ggplot(fixed_plot,
       aes(position, fixed_diffs)) +

  geom_line(linewidth = 1) +

  theme_bw() +

  labs(
    x = "Mitogenome position",
    y = "Fixed differences"
  )
```

```{r}
mean(fixed_df$gene == "control_region")
```


Now I want to see whether there is a similar story with the other groups, not just widespread vs restricted within Australia.
```{r}
uae_ids <- c(
  "628132",
  "628054",
  "628066"
)

other_ids <- c(
  "DRR251525",
  "682064",
  "628131",
  "681968"
)

coding_positions <- 1:15431

dna_matrix <- as.character(mt_DNA)[, coding_positions]

group_list <- list(
  widespread = widespread_ids,
  restricted = restricted_ids,
  uae = uae_ids,
  other = other_ids
)

diagnostic_sites <- list()

for(pos in 1:ncol(dna_matrix)) {

  group_consensus <- c()

  valid_site <- TRUE

  for(g in names(group_list)) {

    ids <- group_list[[g]]

    ids <- ids[ids %in% rownames(dna_matrix)]

    bases <- unique(
      dna_matrix[ids, pos]
    )

    bases <- setdiff(
      bases,
      c("-", "N", "n", "?")
    )

    if(length(bases) != 1) {

      valid_site <- FALSE
      break
    }

    group_consensus[g] <- bases
  }

  if(valid_site &&
     length(unique(group_consensus)) > 1) {

    diagnostic_sites[[length(diagnostic_sites)+1]] <-
      data.frame(
        position = pos,
        widespread = group_consensus["widespread"],
        restricted = group_consensus["restricted"],
        uae = group_consensus["uae"],
        other = group_consensus["other"]
      )
  }
}

diagnostic_df <- do.call(rbind,
                         diagnostic_sites)
```


```{r}
diagnostic_df$gene <- NA

for(i in 1:nrow(bed)) {

  idx <- diagnostic_df$position >= bed$start[i] &
         diagnostic_df$position <= bed$end[i]

  diagnostic_df$gene[idx] <- bed$gene[i]
}

diagnostic_summary <- diagnostic_df %>%
  group_by(gene) %>%
  summarise(n_diag = n()) %>%
  arrange(desc(n_diag))

diagnostic_summary
```

```{r}
group_list <- lapply(group_list, function(ids) {

  intersect(ids,
            rownames(dna_matrix))
})

sapply(group_list, length)


private_sites <- list()


for(target_group in names(group_list)) {

  target_ids <- group_list[[target_group]]

  # Skip empty groups
  if(length(target_ids) == 0) {
    next
  }

  private_positions <- c()

  for(pos in 1:ncol(dna_matrix)) {

    # TARGET BASES
    target_bases <- unique(
      dna_matrix[target_ids, pos, drop = TRUE]
    )

    target_bases <- setdiff(
      target_bases,
      c("-", "N", "n", "?")
    )

    # Must be fixed within lineage
    if(length(target_bases) != 1) {
      next
    }

    target_base <- target_bases[1]

    # OTHER IDS
    other_ids <- unlist(
      group_list[
        names(group_list) != target_group
      ]
    )

    other_ids <- intersect(
      other_ids,
      rownames(dna_matrix)
    )

    # Skip if no comparison samples
    if(length(other_ids) == 0) {
      next
    }

    other_bases <- unique(
      dna_matrix[other_ids, pos, drop = TRUE]
    )

    other_bases <- setdiff(
      other_bases,
      c("-", "N", "n", "?")
    )

    # PRIVATE SNPs
    if(!(target_base %in% other_bases)) {

      private_positions <- c(
        private_positions,
        pos
      )
    }
  }

  private_sites[[target_group]] <- private_positions
}

private_summary <- data.frame(
  lineage = names(private_sites),
  private_diagnostic_snps =
    sapply(private_sites, length)
)

private_summary

private_df <- data.frame(
  lineage = rep(
    names(private_sites),
    sapply(private_sites, length)
  ),
  position = unlist(private_sites)
)

uae_df <- private_df %>%
  filter(lineage == "uae")

ggplot(uae_df,
       aes(position)) +

  geom_histogram(binwidth = 200) +

  theme_bw() +

  labs(
    x = "Mitogenome position",
    y = "Private UAE SNPs"
  )
```
```{r}
diag_positions <- diagnostic_df$position

window_size <- 100
step_size <- 10

windows <- seq(
  1,
  ncol(dna_matrix) - window_size,
  by = step_size
)

diag_counts <- sapply(windows, function(w) {

  sum(diag_positions >= w &
      diag_positions < (w + window_size))
})

mini_windows <- data.frame(
  start = windows,
  end = windows + window_size,
  diagnostic_snps = diag_counts
)

mini_windows %>%
  arrange(desc(diagnostic_snps)) %>%
  head(20)
```


```{r}
inspect_region <- function(start_pos,
                           end_pos,
                           dna_matrix,
                           group_list) {

  get_consensus <- function(ids) {

    apply(
      dna_matrix[ids,
                 start_pos:end_pos,
                 drop = FALSE],
      2,
      function(x) {

        x <- x[!x %in%
                 c("-", "N", "n", "?")]

        if(length(x) == 0) {
          return(NA)
        }

        names(sort(table(x),
                   decreasing = TRUE))[1]
      }
    )
  }

  consensus_df <- data.frame(
    position = start_pos:end_pos
  )

  for(g in names(group_list)) {

    ids <- intersect(
      group_list[[g]],
      rownames(dna_matrix)
    )

    consensus_df[[g]] <- get_consensus(ids)
  }

  consensus_df %>%

    rowwise() %>%

    mutate(
      n_unique = length(
        unique(
          na.omit(
            c_across(-position)
          )
        )
      )
    ) %>%

    ungroup() %>%

    filter(n_unique > 1)
}
```


```{r}
nd5_region <- inspect_region(
  start_pos = 12041,
  end_pos = 12171,
  dna_matrix = dna_matrix,
  group_list = group_list
)

nd5_region
```
```{r}
nd4_region <- inspect_region(
  start_pos = 10601,
  end_pos = 10751,
  dna_matrix = dna_matrix,
  group_list = group_list
)

nd4_region
```



Now I will map them!

```{r}
map_df <- dugong_meta %>%
  mutate(
    decimal_longitude_private = as.numeric(decimal_longitude_private),
    decimal_latitude_private  = as.numeric(decimal_latitude_private),
    plot_location = case_when(
      country == "Australia" ~ location_text,
      country == "UAE" ~ "UAE",
      country == "Palau" ~ "Palau",
      country == "Malaysia" ~ "Malaysia",
      country == "Japan" ~ "Japan",
      country == "New_Caledonia" ~ "New Caledonia",
      TRUE ~ NA_character_
    ),
    shape_group = ifelse(country == "Australia", "Australia", "Outside Australia")
  ) %>%
  filter(!is.na(plot_location))

world <- map_data("world")

map_df_filtered <- map_df %>%
  dplyr::filter(plot_location != "Arcadia Vale")

outside_aus <- c("UAE", "Palau", "Malaysia", "Japan", "New Caledonia")
legend_shapes <- ifelse(
  all_locations_order %in% outside_aus,
  15,  # square
  16   # circle
)

ggplot() +
  borders("world", colour = "grey70", fill = "grey95") +

  geom_point(
    data = map_df_filtered,
    aes(
      x = decimal_longitude_private,
      y = decimal_latitude_private,
      colour = plot_location,
      shape  = shape_group
    ),
    size = 3
  ) +

  scale_colour_manual(
    values = all_location_cols[all_locations_order], 
    breaks = all_locations_order,
    drop = FALSE
  ) +

  scale_shape_manual(
    values = c("Australia" = 16, "Outside Australia" = 15),
    guide = "none"  
  ) +

  coord_sf(
    xlim = c(45, 180),
    ylim = c(-40, 40),
    expand = FALSE
  ) +

  theme_classic() +
  labs(
    x = "Longitude",
    y = "Latitude",
    colour = "Location"
  )
```
And now a map of widespread vs restricted.

```{r}
samples_widespread <- pcoa_df_full %>%
  filter(PC1 < -40) %>%
  mutate(lineage = "Widespread")

samples_restricted <- pcoa_df_full %>%
  filter(PC1 > 40) %>%
  mutate(lineage = "Restricted")

lineage_df <- bind_rows(
  samples_widespread,
  samples_restricted
)
```

```{r}
map_lineages <- dugong_meta %>%
  inner_join(
    lineage_df %>%
      select(bioplatforms_library_id, lineage),
    by = "bioplatforms_library_id"
  ) %>%
  mutate(
    decimal_longitude_private =
      as.numeric(decimal_longitude_private),

    decimal_latitude_private =
      as.numeric(decimal_latitude_private)
  ) %>%
  filter(country == "Australia")
```

```{r}
ggplot() +

  borders("world",
          colour = "grey70",
          fill = "grey95") +

  geom_point(
  data = map_lineages,
  aes(
    x = decimal_longitude_private,
    y = decimal_latitude_private,
    colour = lineage
  ),
  position = position_jitter(width = 0.3, height = 0.3),
  size = 3
) +

  scale_colour_manual(
    values = c(
      "Restricted" = "#D55E00",
      "Widespread" = "#0072B2"
    )
  ) +

  coord_sf(
    xlim = c(110, 155),
    ylim = c(-40, -8),
    expand = FALSE
  ) +

  theme_classic() +

  labs(
    x = "Longitude",
    y = "Latitude",
    colour = "Mitochondrial lineage"
  )

lineage_summary <- map_lineages %>%
  group_by(location_text, lineage) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = lineage,
    values_from = n,
    values_fill = 0
  ) %>%
  arrange(location_text)

lineage_summary
```

Here are the stats for New Caledonia and UAE.

```{r}
new_cal <- read.dna("02_Mitogenomes_local/New_Cal_mtDNA_alignment.fasta", format = "fasta")

# Haplotype diversity
hap.div(new_cal)

# Nucleotide diversity
nuc.div(new_cal)

# Number of haplotypes
haps_new_cal <- haplotype(new_cal)
nrow(haps_new_cal)

#Tajima's D
tajima.test(new_cal)

uae <- read.dna("02_Mitogenomes_local/Abu_Dhabi_mtDNA_aligned.fasta", format = "fasta")

# Haplotype diversity
hap.div(uae)

# Nucleotide diversity
nuc.div(uae)

# Number of haplotypes
haps_uae <- haplotype(uae)
nrow(haps_uae)

#Tajima's D
tajima.test(uae)
```

SKYLINE PLOTS!

```{r, warning=FALSE}
skyline <- read_tsv(
  "02_Mitogenomes_local/86_samples_restricted_group_skyline_plot.txt",
  skip = 1,
  col_types = cols()
)

ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "grey80", alpha = 0.6) +
  geom_line(aes(y = mean), linewidth = 1, colour = "black") +
  geom_line(aes(y = median), linewidth = 0.8, linetype = "dashed") +
  scale_x_reverse() +
  theme_bw()
```


And the sea level data:

```{r, warning=FALSE}
sea_level <- read_tsv(
  "02_Mitogenomes_local/SPRATT2016_SEALEVEL-data.txt",
  comment = "#",
  col_types = cols()
)

sea_level_clean <- sea_level %>%
  mutate(time = age_calkaBP * 1000)

x_limits <- c(41927.65, 0)

p_sea <- ggplot(sea_level_clean, aes(x = time, y = SeaLev_shortPC1)) +
  geom_ribbon(
    aes(
      ymin = SeaLev_shortPC1_err_lo,
      ymax = SeaLev_shortPC1_err_up
    ),
    fill = "steelblue",
    alpha = 0.3
  ) +
  geom_line(linewidth = 1) +
  scale_x_reverse(
    expand = c(0, 0),
    labels = scales::label_number(big.mark = ",")
  ) +
  coord_cartesian(xlim = x_limits) +
  theme_bw() +
  labs(
    x = "Time before present (years)",
    y = "Sea level (m)"
  )

p_sea

sea_level_clean <- sea_level %>%
  mutate(time = age_calkaBP * 1000)

x_limits <- c(41927.65, 0)

p_sea
```

```{r}
skyline_r <- read_tsv(
  "02_Mitogenomes_local/restricted_group_70m_data.txt",
  skip = 1,
  col_types = cols()
) %>%
  filter(!is.na(mean), !is.na(time))

skyline_w <- read_tsv(
  "02_Mitogenomes_local/final_widespread_group_data.txt",
  skip = 1,
  col_types = cols()
) %>%
  filter(!is.na(mean), !is.na(time))

x_limits <- c(41927.65, 0)

y_max <- max(skyline_w$upper, na.rm = TRUE)

x_cutoff <- 40000

x_scale <- scale_x_reverse(expand = c(0, 0))
```

```{r, warning=FALSE}
p_widespread <- ggplot(skyline_w, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "grey80", alpha = 0.7) +
  geom_line(aes(y = mean), linewidth = 1) +
  x_scale +
  coord_cartesian(xlim = c(x_cutoff, 0)) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = " m", accuracy = 0.1)
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = expression("Effective population size (" * N[e] * ")"),
       x = NULL) +
  annotate("text",
           x = x_cutoff * 0.97, y = Inf,
           label = "Clade A",
           hjust = 0, vjust = 1.9,
           size = 4, fontface = "bold")


restricted_y_max <- 2.5e6

p_restricted <- ggplot(skyline_r, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "grey80", alpha = 0.7) +
  geom_line(aes(y = mean), linewidth = 1) +
  x_scale +
  coord_cartesian(xlim = c(x_cutoff, 0)) +
  scale_y_continuous(
    limits = c(0, restricted_y_max),
    labels = scales::label_number(scale = 1e-6, suffix = " m", accuracy = 0.1)
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = expression("Effective population size (" * N[e] * ")"),
       x = NULL) +
  annotate("text",
           x = x_cutoff * 0.97, y = Inf,
           label = "Clade C",
           hjust = 0, vjust = 1.9,
           size = 4, fontface = "bold")



plot_grid(
  p_widespread,
  p_restricted,
  p_sea,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 0.6)
)
```


Making a list of all the samples I used.

```{r}
pcoa_df_joined <- pcoa_df_full %>%
  inner_join(
    dugong_meta %>%
      select(bioplatforms_library_id, specimen_id...6, collector, collection_date, location_text, country, state_or_region),
    by = "bioplatforms_library_id"
  )

write.csv(pcoa_df_joined, "dugong_samples_for_mt_paper.csv", row.names = FALSE)
```

And now a list of all the samples for the haplotype network too.

```{r}
extract_gb_record <- function(record) {

  isolate   <- str_match(record, '/isolate="([^"]+)"')[,2]
  geo       <- str_match(record, '/geo_loc_name="([^"]+)"')[,2]
  country   <- str_match(record, '/country="([^"]+)"')[,2]
  organism  <- str_match(record, '/organism="([^"]+)"')[,2]
  accession <- str_match(record, 'ACCESSION\\s+([A-Z0-9_]+)')[,2]
  version   <- str_match(record, 'VERSION\\s+([A-Z0-9_.]+)')[,2]

  authors <- str_match(
  record,
  "(?m)^\\s*AUTHORS\\s+(.+)$"
  )[,2]

  region <- coalesce(
    geo[!is.na(geo)][1],
    country[!is.na(country)][1],
    NA_character_
  )

  if (!is.na(isolate[1]) &&
      str_detect(isolate[1], "^DU") &&
      (is.na(region) || region == "")) {

    region <- "Thailand"
  }

  tibble(
    TreeID    = version[!is.na(version)][1],
    ID        = isolate[!is.na(isolate)][1],
    Region    = region,
    Organism  = organism[!is.na(organism)][1],
    Accession = accession[!is.na(accession)][1],
    Authors = authors[1],
    Source    = "NCBI"
  )
}

ncbi_meta <- purrr::map_dfr(records, extract_gb_record)
```


```{r}
traits_from_master <- dugong_meta %>%
  transmute(
    TreeID    = bioplatforms_library_id,
    ID        = specimen_id...6,
    Region    = location_text,
    Organism  = "Dugong dugon",
    Accession = NA_character_,
    Authors   = NA_character_,
    Source    = "Local"
  ) %>%
  filter(!is.na(TreeID))
```

```{r}
all_meta_paper <- bind_rows(
  traits_from_master,
  ncbi_meta
) %>%
  distinct(TreeID, .keep_all = TRUE)
```


```{r}
traits_from_master <- dugong_meta %>%
  transmute(
    TreeID          = bioplatforms_library_id,
    ID              = specimen_id...6,
    Region          = location_text,
    Country         = country,
    State           = state_or_region,
    Collector       = collector,
    Collection_date = collection_date,
    Organism        = "Dugong dugon",
    Accession       = NA_character_,
    Authors         = NA_character_,
    Source          = "Local"
  ) %>%
  filter(!is.na(TreeID))
```

```{r}
ncbi_meta <- purrr::map_dfr(records, extract_gb_record) %>%
  mutate(
    Country         = NA_character_,
    State           = NA_character_,
    Collector       = NA_character_,
    Collection_date = NA_character_
  )
```

```{r}
all_meta_paper <- bind_rows(
  traits_from_master,
  ncbi_meta
) %>%
  distinct(TreeID, .keep_all = TRUE)
```

```{r}
write.xlsx(
  all_meta_paper,
  "dugong_mtDNA_paper_sample_metadata.xlsx",
  overwrite = TRUE
)
```

```{r}
clade_A <- scan(
  "02_Mitogenomes_local/Clade_A_samples.txt",
  what = character()
)

clade_C <- scan(
  "02_Mitogenomes_local/Clade_C_samples.txt",
  what = character()
)
```

```{r}
qld_clades <- traits_for_popart %>%
  mutate(
    Clade = case_when(
      ID %in% clade_A ~ "A",
      ID %in% clade_C ~ "C",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    Region == "QLD",
    !is.na(Clade)
  )
```

```{r}
qld_A_seqs <- haplotype_network_seqs[
  rownames(haplotype_network_seqs) %in%
    qld_clades$ID[qld_clades$Clade == "A"],
]

qld_C_seqs <- haplotype_network_seqs[
  rownames(haplotype_network_seqs) %in%
    qld_clades$ID[qld_clades$Clade == "C"],
]
```

Running stats on clade a and c separately.

```{r}
# Queensland sample IDs
qld_ids <- traits_for_popart %>%
  filter(Region == "QLD") %>%
  pull(ID)

# Queensland Clade A IDs
qld_A_ids <- intersect(qld_ids, clade_A)

# Queensland Clade C IDs
qld_C_ids <- intersect(qld_ids, clade_C)

dna_qld_A <- haplotype_network_seqs[
  rownames(haplotype_network_seqs) %in% qld_A_ids,
]

dna_qld_C <- haplotype_network_seqs[
  rownames(haplotype_network_seqs) %in% qld_C_ids,
]

nrow(dna_qld_A)
nrow(dna_qld_C)

stats_qld_A <- calc_stats(dna_qld_A)
stats_qld_C <- calc_stats(dna_qld_C)

stats_qld_A
stats_qld_C
```

Which are the samples that cluster with clades A and C?


```{r}
clade_members <- traits_for_popart %>%
  mutate(
    Clade = case_when(
      ID %in% clade_A ~ "A",
      ID %in% clade_C ~ "C",
      TRUE ~ NA_character_
    )
  )

clade_members %>%
  filter(
    !is.na(Clade),
    !Region %in% c("QLD", "WA", "NT", "NSW", "Australia")
  ) %>%
  arrange(Clade, Region)
```

Mostly Plön et al. samples, which are from ancient DNA.
